Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-05 ind_files/02_BD_processes.tex
Correct minor typos

We assume that the birth and death rate functions have the functional form
\begin{equation}
\label{nD_functional_forms_for_replicator}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= x_ib^{\textrm{(ind)}}_{i}(\mathbf{x}) + \mu Q_i(\mathbf{x})\\
d^{(K)}_i(\mathbf{x}) &= x_id^{\textrm{(ind)}}_i(\mathbf{x})
\end{aligned}
\end{equation}
where $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ and $d^{\textrm{(ind)}}_{i}(\mathbf{x})$ are non-negative functions that respectively describe the per-capita birth and death rate of type $i$ individuals, $\mu \geq 0$ is a constant describing the mutation rate in the population, and $Q_i(\mathbf{x})$ is a non-negative function that describes the additional birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$ that cannot be captured in the per-capita birth rate\footnote{When $x_i = 0$, \emph{i.e.} there are no type $i$ individuals in the population, individuals of type $i$ may still be born through mutations during births of the other types. This cannot be captured in $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ because the term $x_ib^{\textrm{(ind)}}_{i}(\mathbf{x})$ vanishes when $x_i = 0$. Note that no analogous problem exists for the death rate, since the death rate of type $i$ individuals must be 0 when $x_i$ is 0 to ensure that we never have negative population densities.}. Our assumptions of the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates for each type. We define the \emph{Malthusian fitness} of the $i\textsuperscript{th}$ type as $w_i(\mathbf{x}) \coloneqq b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}(\mathbf{x})$, and the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type as $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + b^{\textrm{(ind)}}_{i}(\mathbf{x})$.  The quantity $w_i(\mathbf{x})$ describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$ discounting mutation. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but we will stick to $w_i$ and `fitness' here. It is notable that both $w_i$ and $\tau_i$ depend on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and not just on the density of the focal type. Thus, in general, both the fitness and the turnover rate in our model are frequency-dependent.\\

>>>>

We assume that the birth and death rate functions have the functional form
\begin{equation}
\label{nD_functional_forms_for_replicator}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= x_ib^{\textrm{(ind)}}_{i}(\mathbf{x}) + \mu Q_i(\mathbf{x})\\
d^{(K)}_i(\mathbf{x}) &= x_id^{\textrm{(ind)}}_i(\mathbf{x})
\end{aligned}
\end{equation}
where $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ and $d^{\textrm{(ind)}}_{i}(\mathbf{x})$ are non-negative functions that respectively describe the per-capita birth and death rate of type $i$ individuals, $\mu \geq 0$ is a constant describing the mutation rate in the population, and $Q_i(\mathbf{x})$ is a non-negative function that describes the additional birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$ that cannot be captured in the per-capita birth rate\footnote{When $x_i = 0$, \emph{i.e.} there are no type $i$ individuals in the population, individuals of type $i$ may still be born through mutations during births of the other types. This cannot be captured in $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ because the term $x_ib^{\textrm{(ind)}}_{i}(\mathbf{x})$ vanishes when $x_i = 0$. Note that no analogous problem exists for the death rate, since the death rate of type $i$ individuals must be 0 when $x_i$ is 0 to ensure that we never have negative population densities.}. Our assumptions of the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates for each type. We define the \emph{Malthusian fitness} of the $i\textsuperscript{th}$ type as $w_i(\mathbf{x}) \coloneqq b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}(\mathbf{x})$, and the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type as $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + d^{\textrm{(ind)}}_{i}(\mathbf{x})$.  The quantity $w_i(\mathbf{x})$ describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$ discounting mutation. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but we will stick to $w_i$ and `fitness' here. It is notable that both $w_i$ and $\tau_i$ depend on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and not just on the density of the focal type. Thus, in general, both the fitness and the turnover rate in our model are frequency-dependent.\\

### 2023-03-05 ind_files/04_unification_discussion.tex
Correct minor typos

Stochastic finite population models often exhibit behaviors that are markedly different from their deterministic limits. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic first-principles models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems \citep{black_stochastic_2012,schreiber_does_2022,hastings_transients_2004,shoemaker_integrating_2020}. In the context of eco-evolutionary theory, \cite{doebeli_towards_2017} have called for a reformulation of evolutionary dynamics starting from stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic, a view that has also been echoed by theorists before \citep{metcalf_why_2007,geritz_mathematical_2012}. In this thesis, I present and derive from first principles some fundamental relations (equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance}) that any such reformulation must satisfy. Further, these relations deal with biologically important quantities, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of \cite{queller_fundamental_2017}), or `unifying principles' (in the sense of \cite{lion_theoretical_2018}) for the dynamics of finite populations. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). Just like in infinite population models, the various terms of equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} lend themselves to simple biological interpretation. Note that actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that high dimensional evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos \citep{doebeli_diversity_2017}. We have also neglected any potential complications introduced by genotype-phenotype maps and other such genetic factors, since they can be absorbed into the per-capita birth and death rates. Nevertheless, our equations will still hold for arbitrarily complicated birth and death rates. Like the Price equation, the utility of these equations lies not in their solutions, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations \citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general approach of working with `high-level' processes without specifying system-specific details is part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature  \citep{grafen_formal_2014, queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022}. Such model-independent descriptions that `abstract away' system-specific details almost inevitably come at the cost of precision \citep{levins_strategy_1966, potochnik_idealization_2018}, and are thus intended to complement rather than supersede specific models of specific systems. In our case, our framework complements the various specific studies of social evolution in specific models \citep{constable_demographic_2016, houchmandzadeh_selection_2012, chotibut_evolutionary_2015, mcleod_social_2019} that pointed to noise-induced selection reversing the direction of evolution predicted by deterministic models.\\
Noise-induced selection due to variation in per-capita turnover rate $\tau$ is an evolutionary force only seen in finite populations. Mathematically, this term arises essentially due to It\^o's formula adding an extra term when conducting a change of variables in going from type densities to type frequencies. Biologically, it arises due to different types experiencing a different number of stochastic events (birth and death) in a given time interval. The effects of noise-induced selection are increasingly being recognized in various specific models \citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, parsons_pathogen_2018}, and, recently, have also been shown in some more general settings \citep{week_white_2021,mazzolini_universality_2022,kuosmanen_turnover_2022}. This force has some important practical implications. For one, simulation studies that work with evolutionary individual-based or agent-based models should be careful about whether interactions effects are incorporated into birth rates or death rates, since this seemingly arbitrary choice can have unintended consequences due to noise-induced selection \citep{mcleod_social_2019,kuosmanen_turnover_2022}. Being mindful of noise-induced selection is also important for applied fields like conservation and population management that regularly deal with small populations (\emph{i.e.} populations which likely experience stronger noise-induced selection). For example, when trying to increase the population of a hypothetical desired species in a multispecies community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$ - Decreasing the death rate leads to a decrease in $\tau_i$, which leads to positive noise-induced selection (from \eqref{nD_stochastic_RM}), whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community. If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that our actions \emph{increased} the growth rate of this species. Similarly, when trying to infer or predict the future trajectory of the relative abundance of a species (or phenotype, allele, etc.) from empirical data, our equations indicate that measuring the growth rate of populations is not, in general, sufficient for accurate prediction even in completely controlled environments. The growth rate $w_i = b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}$ of a species $i$ only specifies the difference between its per-capita birth and death rates, whereas the complete stochastic dynamics also depend on the total turnover $\tau_i = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + d^{\textrm{(ind)}}_{i}$ (\emph{i.e.} the sum of the per-capita birth and death rates). Two species can have very different $\tau_i$ while having the same $w_i$, leading to deviations from predictions that only account for growth rates in smaller communities due to noise-induced selection causing a systematic deviation from true neutrality. Thus, equations \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price} imply that for evolution to be \emph{truly} neutral in finite populations, it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. A similar phenomenon has also been recognized in some finite population ecological models, where seemingly neutral models (equal growth rate) nevertheless select for individuals with greater longevity \citep{lin_features_2012, oliveira_advantage_2017} \\

>>>>

Stochastic finite population models often exhibit behaviors that are markedly different from their deterministic limits. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic first-principles models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems \citep{black_stochastic_2012,schreiber_does_2022,hastings_transients_2004,shoemaker_integrating_2020}. In the context of eco-evolutionary theory, \cite{doebeli_towards_2017} have called for a reformulation of evolutionary dynamics starting from stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic, a view that has also been echoed by theorists before \citep{metcalf_why_2007,geritz_mathematical_2012}. In this thesis, I present and derive from first principles some fundamental relations (equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance}) that any such reformulation must satisfy. Further, these relations deal with biologically important quantities, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of \cite{queller_fundamental_2017}), or `unifying principles' (in the sense of \cite{lion_theoretical_2018}) for the dynamics of finite populations. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). Just like in infinite population models, the various terms of equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} lend themselves to simple biological interpretation. Note that actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that high dimensional evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos \citep{doebeli_diversity_2017}. We have also neglected any potential complications introduced by genotype-phenotype maps and other such genetic factors, since they can be absorbed into the per-capita birth and death rates. Nevertheless, our equations will still hold for arbitrarily complicated birth and death rates. Like the Price equation, the utility of these equations lies not in their solutions, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations \citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general approach of working with `high-level' processes without specifying system-specific details is part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature  \citep{grafen_formal_2014, queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022}. Such model-independent descriptions that `abstract away' system-specific details almost inevitably come at the cost of precision \citep{levins_strategy_1966, potochnik_idealization_2018}, and are thus intended to complement rather than supersede specific models of specific systems.\\
Noise-induced selection due to variation in per-capita turnover rate $\tau$ is an evolutionary force only seen in finite populations. Mathematically, this term arises essentially due to It\^o's formula adding an extra term when conducting a change of variables in going from type densities to type frequencies. Biologically, it arises due to different types experiencing a different number of stochastic events (birth and death) in a given time interval. The effects of noise-induced selection are increasingly being recognized in various specific models \citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, parsons_pathogen_2018}, and, recently, have also been shown in some more general settings \citep{week_white_2021,mazzolini_universality_2022,kuosmanen_turnover_2022}. This force has some important practical implications. For one, simulation studies that work with evolutionary individual-based or agent-based models should be careful about whether interactions effects are incorporated into birth rates or death rates, since this seemingly arbitrary choice can have unintended consequences due to noise-induced selection \citep{mcleod_social_2019,kuosmanen_turnover_2022}. Being mindful of noise-induced selection is also important for applied fields like conservation and population management that regularly deal with small populations (\emph{i.e.} populations which likely experience stronger noise-induced selection). For example, when trying to increase the population of a hypothetical desired species in a multispecies community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$ - Decreasing the death rate leads to a decrease in $\tau_i$, which leads to positive noise-induced selection (from \eqref{nD_stochastic_RM}), whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community. If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that our actions \emph{increased} the growth rate of this species. Similarly, when trying to infer or predict the future trajectory of the relative abundance of a species (or phenotype, allele, etc.) from empirical data, our equations indicate that measuring the growth rate of populations is not, in general, sufficient for accurate prediction even in completely controlled environments. The growth rate $w_i = b^{\textrm{(ind)}}_{i} - d^{\textrm{(ind)}}_{i}$ of a species $i$ only specifies the difference between its per-capita birth and death rates, whereas the complete stochastic dynamics also depend on the total turnover $\tau_i = b^{\textrm{(ind)}}_{i} + d^{\textrm{(ind)}}_{i}$ (\emph{i.e.} the sum of the per-capita birth and death rates). Two species can have very different $\tau_i$ while having the same $w_i$, leading to discrepancies from predictions that only account for growth rates ($w$) in smaller communities due to noise-induced selection causing a systematic deviation from true neutrality. Thus, equations \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price} imply that for evolution to be \emph{truly} neutral in finite populations, it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. A similar phenomenon has also been recognized in some finite population ecological models, where seemingly neutral models (equal growth rate) nevertheless select for individuals with greater longevity \citep{lin_features_2012, oliveira_advantage_2017} \\
